Do motif roles provide unique information about species risk of extinction?\\ \medskip Supplementary Information

### 2021-10-01 manuscript/supplement_motifs.tex
synced order of appendices with main text

\section{Network creation and processing}

@@

\section{Calculating persistence}

@@

\section{Statistical analysis}

@@

        As with network profiles, species' motif participation may covary with other measures of network structure and species' roles within the network.
        To establish this context for motif participation, we first fit a PERMANOVA~\citep{Anderson2001} to test whether a species' overall motif participation role is related to network size, connectance, or the interaction between them.
        To avoid inflating similarity between species which do not appear in the same motifs (i.e., which have ``shared zeros''), we use Bray-Curtis dissimilarity between motif participation roles.
        Significance was calculated using 9999 permutations.

        PERMANOVA tests may give false positive results if variability is not homogeneous among groups (i.e., levels of network size or connectance).
        To test whether this applies in our case, we fit an ANOVA of the dispersion of motif participation roles about the centroid for each combination of network size and connectance. 
        We then regressed dispersion against persistence to test whether networks with higher or lower persistence had more consistent motif profiles.
        We fit the PERMANOVA using the function `adonis', calculated dispersion using the function `betadisper', and fit the ANOVA using the function `anova', all from the R~\citep{R} package \emph{vegan}~\citep{vegan}.
        We fit the regression using the R~\citep{R} base function `lm'.

        To obtain a more detailed picture of the relationships between motif participation and global network structure, we fit a set of four linear models (one per motif):

@@

    \subsection{How does persistence vary with species and network properties?}

        \subsubsection{How does network mean persistence vary depending on network properties?}

@@

        \subsubsection{How does persistence vary with in-degree and trophic level?}

@@

@@

\section{SB: Motif profiles and global properties}

\section{SC: Persistence and motif profiles}

    \begin{table}[h!]
        \caption{Coefficients ($\beta$) for models relating persistence to the proportion of a network profile, the level of disturbance to basal resources (corresponding to a probability of extinction ranging between 0.1 and 0.5), and their interaction. Predictors were centered and scaled before model fitting. All effects were significant ($p$\textless0.001). We also provide the $R^2$. }
        \label{motif_profile_tab}
        \centering
        \begin{tabular}{c|c c c c c | c}
            Motif & Intercept & Proportion & Disturbance & Interaction & $R^2$\\
            \hline
            Omnivory & 0.572 & -0.0133 & -0.139 & -4.89$\times10^{-4}$ & 0.908 \\
            Apparent competition & 0.572 & 0.0105 & -0.139 & 4.78$\times10^{-4}$ & 0.905 \\
            Direct competition & 0.572 & 2.62$\times10^{-3}$ & -0.139 & 3.84$\times10^{-4}$ & 0.901 \\
            Three-species chain & 0.572 & 6.12$\times10^{-3}$ & -0.139 & -1.90$\times10^{-4}$ & 0.901 \\ 
        \end{tabular}
    \end{table}

\section{SD: Persistence and motif participation}

\section{SE: Motif-persistence relationships \& network size}

\section{SF: Persistence, degree, and trophic level}

\section{SG: Motif participation and network properties}

    \begin{figure}[h!]
        \centering
        \includegraphics[height=.65\textheight]{figures/participation_vs_SC.eps}
        \caption{The proportion of each motif in a consumer's participation vector varied with network size, connectance, and their interaction. These relationships were steepest for apparent competition and omnivory.}
        \label{fig:roles_vs_SC}
    \end{figure}
\clearpage

@@

    % \section{How do counts of motifs in a network profile vary with S and C?}

    %     As expected, all four motifs had higher counts in larger and more highly connected networks.
    %     The precise effects of size, connectance, and their interaction differed between motifs  (Table~\ref{network_motif_lms}).
    %     This suggests that the meso-scale structure of a network varies with size and connectance, not just the total number of motifs.
    %     Indeed, the motif profile of a network is significantly related to its size, connectance, and their interaction ($F_{1,2996}$=2279, $p$\textless0.001; $F_{1,2996}$=3067, $p$\textless0.001; and $F_{1,2996}$=1187, $p$\textless0.001, respectively).

    %     \begin{table}[h!]
    %         \centering
    %         \caption{Coefficients for linear regressions relating the total count of all motifs or the count of each three-species motif in a network to network size, connectance, and their interaction. Coefficients in \textbf{bold} are significant (\textless0.05). Note that, while there are 13 unique three-species motifs, only four are possible in acyclic networks such as those considered here.}
    %        \label{network_motif_lms}
    %        \begin{tabular}{c|c c c c c}
    %         Motif & Intercept & Size & Connectance & Interaction \\
    %         \hline
    %         Total count & \textbf{-2.85$\times10^3$} & \textbf{46.6} & \textbf{-2.12$\times10^{5}$} & \textbf{4.72$\times10^{3}$} \\
    %         \hline
    %         Thee-species Chain & \textbf{-1.40$\times10^{3}$} & \textbf{26.1} & \textbf{-3.74$\times10^{4}$} & \textbf{8.58$\times10^{2}$} \\
    %         Apparent Competition & \textbf{-2.23$\times10^3$} & \textbf{42.0} & 
    %         \textbf{-7.26$\times10^4$} & \textbf{1.63$\times10^3$} \\
    %         Direct Competition & 26.6 & 0.777 & \textbf{-4.85$\times10^4$} & \textbf{9.89$\times10^2$} \\
    %         Omnivory & \textbf{7.57} & \textbf{-22.2} & \textbf{-5.37$\times10^4$} & \textbf{1.24$\times10^3$} \\
    %         \hline
    %         \end{tabular}
    %     \end{table}

    %     \begin{figure}[h!]
    %         \centering
    %         \includegraphics[width=.75\textwidth]{figures/motif_profile_dispersion.pdf}
    %         \caption{Dispersion of motif profiles within a network about the centroid for that combination of network size and connectance. Each circle represents a single network; circle colour indicates network size. Scatter has been introduced about each level of connectance to increase visual clarity; the levels of connectance we consider are indicated by vertical black lines.}
    %         \label{dispersion_rawmotifs}
    %     \end{figure}

    %     Motif profiles were similarly variable across networks with different sizes, but were not homogeneously variable across connectances or size-connectance combinations ($F_{5,2994}$=0.365, $p$=0.873; $F_{4,2995}$=4.49, $p$=0.001; and $F_{29,2970}$=27.1, $p$\textless0.001 , respectively).
    %     These differences in variability of motif profiles can cause false positives in the PERMANOVA.
    %     Variability in motif profiles was lowest among large networks with moderate connectance and was highest among the least-connected and most-connected networks we considered (Fig.~\ref{dispersion_rawmotifs}).

    %     \begin{figure}[h!]
    %         \centering
    %         \includegraphics[width=\textwidth]{figures/absolute_lmer_allCS.pdf}
    %         \caption{The effect of number of times a species appears (x-axis) in any of the various motifs (columns) on persistence (y-axis). The effect of each motif participation is fitted by linear mixed-effect models, following Equation~\ref{absreq}, except that the disturbance levels are fitted separately. The different colored lines indicate various disturbances on the basal level, from $\pi_{disturbed} = 0.1$ (in the top) to $\pi_{disturbed} = 0.5$ (in the bottom).}
    %         \label{fig:abs_lmer_all}
    %     \end{figure}

@@

        \begin{table}[h!]
            \caption{Coefficients ($\beta$) and $p$-values for linear regressions of the frequency of each motif position against the proportion of the network made up of basal resources (pBasal), connectance, and their interaction.}
            \label{BC_tab}
            \footnotesize
            \begin{tabular}{c c | c c | cc | cc |}
            Motif & Position & \multicolumn{2}{c}{pBasal} & \multicolumn{2}{c}{Connectance} & \multicolumn{2}{c}{Interaction} \\
            & & $\beta$ & $p$ & $\beta$ & $p$ & $\beta$ & $p$ \\
            \hline
            Chain & bottom & -0.116 & \textless0.001 & -0.266 & \textless0.001 & 0.217 & \textless0.001 \\
            Chain & middle & 0.0779 & \textless0.001 & -0.107 & \textless0.001 & -0.386 & \textless0.001 \\
            Chain & top & 0.0829 & \textless0.001 & 0.0776 & \textless0.001 & -0.356 & \textless0.001 \\
            \hline
            Omnivory & bottom & -0.104 & \textless0.001 & 0.203 & \textless0.001 & 0.937 & \textless0.001 \\
            Omnivory & middle & -0.0668 & \textless0.001 & 0.217 & \textless0.001 & 1.24 & \textless0.001 \\
            Omnivory & top & -0.0269 & \textless0.001 & 0.281 & \textless0.001 & 0.638 & \textless0.001 \\
            \hline
            App. comp. & bottom & -0.237 & \textless0.001 & -0.874 & \textless0.001 & 1.76 & \textless0.001 \\
            App. comp. & top & 0.154 & \textless0.001 & 0.0279 & \textless0.001 & -0.0947 & 0.013 \\
            \hline
            Dir. comp. & bottom & -0.0631 & \textless0.001 & -0.0247 & \textless0.001 & -0.418 & \textless0.001 \\
            Dir. comp. & top & 0.298 & \textless0.001 & 0.466 & \textless0.001 & =3.53 & \textless0.001 \\
            \hline
            \end{tabular}
            \end{table}

\section{Motif profiles in different network sets}

@@

    \begin{table}[h!]
        \centering
        \caption{Proportions of the four motifs which appear in Bayesian networks (three-species chain: `Chain', apparent competition: `AC', direct competition `DC', and omnivory: `Omni') and other motifs (`Other') in the motif profiles of acyclic, filtered, and original networks. Acyclic networks are those used in the main text analyses. Filtered networks are those used in the main text analyses \emph{before} being rendered acyclic. Original networks are all simulated networks, including those removed because of overly high maximum trophic levels.}
        \label{tab:acyclic_proportions}            \footnotesize
        \begin{tabular}{c|l l l | l l l | l l l |}
        & \multicolumn{3}{c|}{Acyclic networks} & \multicolumn{3}{c|}{Filtered networks} & \multicolumn{3}{c|}{Original networks} \\
        Motif & Mean (SD) & Min & Max & Mean (SD) & Min & Max & Mean (SD) & Min & Max \\
        \hline
        Chain & 0.235 (0.042) & 0.109 & 0.412 & 0.225 (0.045) & 0.100 & 0.412 & 0.218 (0.043) & 0.096 & 0.413 \\
        AC & 0.422 (0.064) & 0.250 & 0.765 &
        0.417 (0.067) & 0.233 & 0.765 & 0.397 (0.064) & 0.231 & 0.765 \\
        DC & 0.179 (0.041) & 0.050 & 0.359 & 0.155 (0.046) & 0.040 & 0.359 & 0.140 (0.046) & 0.031 & 0.359 \\
        Omni & 0.164 (0.079) & 0.006 & 0.346 & 0.156 (0.075) & 0.006 & 0.319 & 0.184 (0.073) & 0.006 & 0.329 \\
        \hline
        Other & \multicolumn{3}{c|}{NA} & 0.047 (0.042) & 0.00 & 0.200 &
        0.062 (0.046) & 0.00 & 0.258 \\
        \end{tabular}
    \end{table}

% \section{Persistence and count-based motif participation}

%     \subsection*{Methods}

%         We test whether participating in higher absolute numbers (i.e., counts) of a motif increases persistence using a set of models that parallels those used to test whether having a greater proportion of the role made up by a particular motif (see \emph{Main Text}).  
%         Specifically, we fit four linear mixed-effect models (LMMs; one per motif included in the Bayesian networks).
%         In each case, we considered only non-basal species (i.e., those with at least one prey) as the persistence of basal species is determined only by their baseline probability.
%         These models included the absolute number of times the focal species appeared in the focal motif, the increase in probability of extinction for basal species, and the interaction between the two as well as a random intercept for the interaction between species richness and connectance (equations.
%         All models were fit using the R~\citep{R} function `lmer' from the package \emph{lmerTest}~\citep{lmerTest}.
%         We calculated marginal (fixed effects only) and conditional (fixed and random effects) $R^2$ using the R~\citep{R} function `r.squaredGLMM' from the package \emph{MuMIn}~\citep{MuMIn}.

%             The formula for these LMMs are:         
%             \begin{equation}
%                 \Psi_{ik} \approx \chi_{i} + \pi_{disturbed_k} + \chi_{i}\pi_{disturbed_k} + S:C_{i} ,
%                 \label{absreq}
%             \end{equation}

%             \noindent where $\Psi_ik$ is the persistence of species $i$ during disturbance level $k$, $\chi_{i}$ is the number of times species $i$ appears in the focal motif, $\pi_{disturbed_k}$ is the probability of extinction for a basal resource in disturbance level $k$, and $S:C_{i}$ is a random intercept for the species richness and connectance of the network containing species $i$.

%             % Additionally, we performed a similar linear regression of persistence against motif participation within each simulated network and for each level of disturbance to basal resources (XX regressions total).
%             % These regressions were fit using the R~\citep{R} base function `lm'.
%             % We then analyse the distribution of slopes for the effect of motif participation across levels of disturbance, network size, and connectance.

%     \subsection*{Results}

%         \begin{figure}[h!]
%             \centering
%             \includegraphics[width=\textwidth]{figures/absolute_lmer_allCS.pdf}
%             \caption{The effect of number of times a species appear (x-axis) in any of the various motifs (columns) on persistence (y-axis). The effect of each motif participation is fitted by linear mixed-effect models, following \emph{Equation 2; Main Text}, except that the disturbance levels are fitted separately. The different colored lines indicate various disturbances on the basal level, from $\pi_{disturbed} = 0.1$ (in the top) to $\pi_{disturbed} = 0.5$ (in the bottom).}
%             \label{fig:abs_lmer_all}
%         \end{figure}

%         Persistence increased significantly with the number of times a species appeared in any of the four possible motifs in the food webs (Table~\ref{tab:absolute_number} and Figure~\ref{fig:abs_lmer_all}). 
%         These relationships were slightly stronger for the number of times a species appears in the three-species chain and direct competition motifs and somewhat weaker for apparent competition and omnivory. The absolute difference between the motifs was small, with a clear positive relationship between persistence and motif count present in all cases. 
%         As expected, persistence decreased significantly with increasing disturbance of basal resources and the strength of this relationship was very similar across the four motifs.
%         The level of disturbance to basal resources also affected the relationship between persistence and motif counts.
%         For all four motifs, there was a negative interaction between motif count and disturbance such that, for high levels of disturbance, participating more often in any of the motifs would result in decreased persistence.

%         \begin{table}[h!]
%             \caption{Coefficients ($\beta$) for models relating persistence to the number of times a species appears in the focal motif, the level of disturbance to basal resources, and their interaction, as well as a random intercept to account for differences in persistence across levels of network size and connectance (Equation~\ref{absreq}). $p$-values for all coefficients in all models were \textless0.001. We also provide the marginal $R^2_M$ (fixed effects only) and conditional $R^2_C$ (fixed and random effects) for each model.}
%             \label{tab:absolute_number}        \centering
%             \begin{tabular}{c|c  c c  c | c c |}
%             Motif & Intercept & Motif count & Disturbance & Interaction & $R^2_M$ & $R^2_C$ \\
%             \hline
%             3-sp Chain & 0.762 & 1.14$\times10^{-4}$ & -0.388 & -1.36$\times10^{-4}$ & 0.897 & 0.913 \\
%             Apparent Compet. & 0.770 & 3.41$\times10^{-5}$ & -0.397 & -4.36$\times10^{-5}$ & 0.898 & 0.911\\
%             Direct Competition & 0.764 & 1.22$\times10^{-4}$ & -0.398 & -8.46$\times10^{-5}$ & 0.895 & 0.914 \\
%             Omnivory & 0.770 & 6.01$\times10^{-5}$ & -0.400 & -5.91$\times10^{-5}$ & 0.897 & 0.912 \\                
%             \end{tabular}
%             \end{table}

% \clearpage

>>>>

\section{Supplemental methods: Network creation}

@@

\section{Supplemental methods: Calculating persistence}

@@

\section{Supplemental methods: Statistical analysis}

@@

        Species' motif participation may covary with other measures of network structure and species' roles within the network.
        To determine the relationships between motif participation and global network structure, we fit a set of four linear models (one per motif):

@@

    \subsection{How does persistence vary with other properties?}

        \subsubsection{Network mean persistence and network properties}

@@

        \subsubsection{Species persistence and species properties}

@@

\section{Results: Persistence and motif participation}

\section{Results: Motif participation and network properties}

    \begin{figure}[h!]
        \centering
        \includegraphics[height=.5\textheight]{figures/participation_vs_SC.eps}
        \caption{The proportion of each motif in a consumer's participation vector varied with network size, connectance, and their interaction. These relationships were steepest for apparent competition and omnivory.}
        \label{fig:roles_vs_SC}
    \end{figure}
\clearpage

\section{Results: Persistence, degree, and trophic level}

\section{Results: Motif-persistence relationships}

\section{Results: Persistence and motif profiles}

    \begin{table}[h!]
        \caption{Coefficients ($\beta$) for models relating persistence to the proportion of a network profile, the level of disturbance to basal resources (corresponding to a probability of extinction ranging between 0.1 and 0.5), and their interaction. Predictors were centered and scaled before model fitting. All effects were significant ($p$\textless0.001). We also provide the $R^2$. }
        \label{motif_profile_tab}
        \centering
        \begin{tabular}{c|c c c c c | c}
            Motif & Intercept & Proportion & Disturbance & Interaction & $R^2$\\
            \hline
            Omnivory & 0.572 & -0.0133 & -0.139 & -4.89$\times10^{-4}$ & 0.908 \\
            Apparent competition & 0.572 & 0.0105 & -0.139 & 4.78$\times10^{-4}$ & 0.905 \\
            Direct competition & 0.572 & 2.62$\times10^{-3}$ & -0.139 & 3.84$\times10^{-4}$ & 0.901 \\
            Three-species chain & 0.572 & 6.12$\times10^{-3}$ & -0.139 & -1.90$\times10^{-4}$ & 0.901 \\ 
        \end{tabular}
    \end{table}
\clearpage

\section{Results: Motif profiles and global properties}

@@

@@

@@

        \begin{table}[h!]
            \caption{Coefficients ($\beta$) and $p$-values for linear regressions of the frequency of each motif position against the proportion of the network made up of basal resources (pBasal), connectance, and their interaction.}
            \label{BC_tab}
            \footnotesize
            \begin{tabular}{c c | c c | cc | cc |}
            Motif & Position & \multicolumn{2}{c}{pBasal} & \multicolumn{2}{c}{Connectance} & \multicolumn{2}{c}{Interaction} \\
            & & $\beta$ & $p$ & $\beta$ & $p$ & $\beta$ & $p$ \\
            \hline
            Chain & bottom & -0.116 & \textless0.001 & -0.266 & \textless0.001 & 0.217 & \textless0.001 \\
            Chain & middle & 0.0779 & \textless0.001 & -0.107 & \textless0.001 & -0.386 & \textless0.001 \\
            Chain & top & 0.0829 & \textless0.001 & 0.0776 & \textless0.001 & -0.356 & \textless0.001 \\
            \hline
            Omnivory & bottom & -0.104 & \textless0.001 & 0.203 & \textless0.001 & 0.937 & \textless0.001 \\
            Omnivory & middle & -0.0668 & \textless0.001 & 0.217 & \textless0.001 & 1.24 & \textless0.001 \\
            Omnivory & top & -0.0269 & \textless0.001 & 0.281 & \textless0.001 & 0.638 & \textless0.001 \\
            \hline
            App. comp. & bottom & -0.237 & \textless0.001 & -0.874 & \textless0.001 & 1.76 & \textless0.001 \\
            App. comp. & top & 0.154 & \textless0.001 & 0.0279 & \textless0.001 & -0.0947 & 0.013 \\
            \hline
            Dir. comp. & bottom & -0.0631 & \textless0.001 & -0.0247 & \textless0.001 & -0.418 & \textless0.001 \\
            Dir. comp. & top & 0.298 & \textless0.001 & 0.466 & \textless0.001 & =3.53 & \textless0.001 \\
            \hline
            \end{tabular}
            \end{table}
\clearpage

\section*{SK: Motif profiles in different network sets}

@@

    \begin{table}[h!]
        \centering
        \caption{Proportions of the four motifs which appear in Bayesian networks (three-species chain: `Chain', apparent competition: `AC', direct competition `DC', and omnivory: `Omni') and other motifs (`Other') in the motif profiles of acyclic, filtered, and original networks. Acyclic networks are those used in the main text analyses. Filtered networks are those used in the main text analyses \emph{before} being rendered acyclic. Original networks are all simulated networks, including those removed because of overly high maximum trophic levels.}
        \label{tab:acyclic_proportions}            \footnotesize
        \begin{tabular}{c|l l l | l l l | l l l |}
        & \multicolumn{3}{c|}{Acyclic networks} & \multicolumn{3}{c|}{Filtered networks} & \multicolumn{3}{c|}{Original networks} \\
        Motif & Mean (SD) & Min & Max & Mean (SD) & Min & Max & Mean (SD) & Min & Max \\
        \hline
        Chain & 0.235 (0.042) & 0.109 & 0.412 & 0.225 (0.045) & 0.100 & 0.412 & 0.218 (0.043) & 0.096 & 0.413 \\
        AC & 0.422 (0.064) & 0.250 & 0.765 &
        0.417 (0.067) & 0.233 & 0.765 & 0.397 (0.064) & 0.231 & 0.765 \\
        DC & 0.179 (0.041) & 0.050 & 0.359 & 0.155 (0.046) & 0.040 & 0.359 & 0.140 (0.046) & 0.031 & 0.359 \\
        Omni & 0.164 (0.079) & 0.006 & 0.346 & 0.156 (0.075) & 0.006 & 0.319 & 0.184 (0.073) & 0.006 & 0.329 \\
        \hline
        Other & \multicolumn{3}{c|}{NA} & 0.047 (0.042) & 0.00 & 0.200 &
        0.062 (0.046) & 0.00 & 0.258 \\
        \end{tabular}
    \end{table}
\clearpage
